---
title: "Phagocytosis Analysis TI22"
author: "Jesse Garrett-Larsen"
date: "2024-01-29"
output: 
  html_document: 
    keep_md: yes
---
Analysis of Phagocytosis assay 

```{r}
rm(list=ls())

library(ggplot2)
library(tidyverse)
library(lme4)
library(effects)
library(AICcmodavg)
library(DHARMa)
library(emmeans)

ti <- read.csv("ti_merged_data.csv")

#select only phagocytosis assay and make new wbc total, phago total, and phago score columns
ti <- ti %>%
  filter(Well.1.WBC != "NA") %>% #select only phagocytosis assay
  mutate(wbc_total=Well.1.WBC+Well.2.WBC+Well.3.WBC+Well.4.WBC)%>% #sum of wbc quadruplicates
  mutate(phago_total=Well.1.Phago+Well.2.Phago+Well.3.Phago+Well.4.Phago) %>% #sum of phago quadruplicates
  mutate(phago_score = phago_total/(wbc_total+phago_total)) #phago score = phagocytic / total

ti <- ti%>%
  filter(dpi != 2) %>% #remove dpi 2 because only a small subset were run day 2
  drop_na(phago_score)
```

What do the data look like?
```{r}
g.phago <- ggplot(data=ti, aes(x=groups, y=phago_score, shape=groups))+
  geom_jitter(width = .1, size=2, stroke=1)+
  stat_summary(fun.y=mean,
               fun.min = function(y) mean(y)-sd(y),
               fun.max = function(y) mean(y)+sd(y),
               geom= "errorbar", size=0.2, width=0.1, alpha=0.75)+ #error bars
  stat_summary(aes(group=groups), fun=mean, alpha=1, size=3, shape="-")+
  labs(x="Treatment", y="Phagocytosis Score", shape="Treatment Groups")+
  scale_shape_manual(values=c(1, 2, 3, 4))

g.phago
```



*Phagocytosis Model*
I use a generalized linear mixed model with a binomial distribution to test whether temperature or MG treatment affected phagocytosis activity of white blood cells.
Each cell is treated as phagocytic or not.

Infection with MG (p = 0.0296), but not temperature treatment (0.161) significantly affects phagocytic activity.
```{r Phagocytosis binomial GLM}

glm1 <- glmer(phago_score~temp+treatment + (1|band_number), weights=wbc_total+phago_total, data=ti, family="binomial")

summary(glm1)

simulateResiduals(glm1, plot = T)

plot(allEffects(glm1))

#Model Selection
p1 <- glmer(phago_score~temp+treatment + (1|band_number), weights=wbc_total+phago_total, data=ti, family="binomial")
p2<- glmer(phago_score~temp*treatment + (1|band_number), weights=wbc_total+phago_total, data=ti, family="binomial")
p3<- glmer(phago_score~temp + (1|band_number), weights=wbc_total+phago_total, data=ti, family="binomial")
p4<- glmer(phago_score~treatment + (1|band_number), weights=wbc_total+phago_total, data=ti, family="binomial")
p5<- glmer(phago_score~1 + (1|band_number), weights=wbc_total+phago_total, data=ti, family="binomial")
p6 <- glmer(phago_score~groups + (1|band_number), weights=wbc_total+phago_total, data=ti, family="binomial")
p7 <- glmer(phago_score~temp+treatment + sex + (1|band_number), weights=wbc_total+phago_total, data=ti, family="binomial")
aictab(cand.set=list(p1, p2, p3, p4, p5, p6, p7), modnames=c("p1", "p2", "p3", "p4", "p5", "p6", "p7"))

#p4 best model, but use p1 to keep both temp and treatment in model
summary(p4)
summary(p1)

resid <- simulateResiduals(p1)
plot(resid)

emm_results <- emmeans(glm1, ~ temp + treatment, scale = "response")
pairs(emm_results)
```

```{r Phagocytosis Beta Regression}
#install.packages("betareg")
library(betareg)

#beta regression
beta_mod <- betareg(phago_score ~ temp + treatment, data=ti)
summary(beta_mod)
plot(allEffects(beta_mod))
plot(resid(beta_mod))

emm_results <- emmeans(beta_mod, ~ temp + treatment, scale = "response")
pairs(emm_results)
```


```{r Phagocytosis t test}
#t test
treat_inf <- ti$phago_score[ti$treatment == "Infected"]
treat_cont <- ti$phago_score[ti$treatment == "Control"]
temp_warm <- ti$phago_score[ti$temp == "Warm"]
temp_cold <- ti$phago_score[ti$temp == "Cold"]
t.inf <- t.test(treat_inf, treat_cont)
t.temp <- t.test(temp_warm, temp_cold)

print(t.inf)
print(t.temp)
```

**I can't seem to get the confidence intervals for my model predictions using the predict() function**
```{r Model Predictions}
#Model Predictions
dat.new=expand.grid(temp=unique(ti$temp),
                   treatment=unique(ti$treatment))#new grid to put predictions into
dat.new$yhat = predict(beta_mod, type="response", newdata=dat.new, re.form=NA) #predicted values based off beta_mod
dat.new$inf_temp <-  paste(dat.new$treatment, dat.new$temp, sep = "_")
head(dat.new)

#plot predicted values over raw data
phago.pred <- ggplot(data=ti, aes(x=treatment, y=phago_score, shape=temp))+
  geom_jitter(size=2, width=0.1)+
  geom_line(data=dat.new, aes(x=treatment, y=yhat, linetype=temp, group=temp), size=1)+ #model predictions
  stat_summary(aes(group=treatment), fun=mean, alpha=1, size=3, shape="-")+
  stat_summary(data=ti, aes(group=treatment), fun.y=mean,
               fun.min = function(y) mean(y)-sd(y),
               fun.max = function(y) mean(y)+sd(y),
               geom= "errorbar", size=0.5, width=0.1, alpha=0.75)+ #error bars
  labs(x="Infection Treatment", y="Phagocytosis Score", linetype="Temperature", shape="Temperature")+
  scale_shape_manual(values=c(0, 15),  labels= c("Cold", "Warm"))+
  scale_linetype_manual(values = c("dotted", "solid"), labels= c("Cold", "Warm"))

phago.pred
```

####Heterogeneity
Subset the phagocytois assay by groups and look at their variance
```{r Binomial model heterogeneity}
glm.phago2 <- glmer(phago_score~1 + (1|band_number), weights=wbc_total+phago_total, data=ti, family="binomial")
glm.phago2a <- glmer(phago_score~1 + (1|band_number), weights=wbc_total+phago_total, data=subset(ti, inf_temp=="MG_ST"), family="binomial")
glm.phago2b <- glmer(phago_score~1 + (1|band_number), weights=wbc_total+phago_total, data=subset(ti, inf_temp=="Sham_ST"), family="binomial")
glm.phago2c <- glmer(phago_score~1 + (1|band_number), weights=wbc_total+phago_total, data=subset(ti, inf_temp=="MG_TN"), family="binomial")
glm.phago2d <- glmer(phago_score~1 + (1|band_number), weights=wbc_total+phago_total, data=subset(ti, inf_temp=="Sham_TN"), family="binomial")

summary(glm.phago2a); summary(glm.phago2b); summary(glm.phago2c); summary(glm.phago2d)

#variance by group from models above
var <- c(0.9908, 0.7769, 0.3096, 0.5205)
sd <- c(0.9954, 0.8814, 0.5564, 0.7215)
g <- c("Infected Cold", "Control Cold", "Infected Warm", "Control Warm")
temp <- c("Cold", "Cold", "Warm", "Warm")

variability <- data.frame(g,var, sd, temp)
variability$min <- variability$var - variability$sd
variability$max <- variability$var + variability$sd

variability$g <- factor(variability$g, levels = c("Control Cold", "Infected Cold", "Control Warm", "Infected Warm"))


g.var <- ggplot(variability, aes(x=g, y=var, shape=g))+
  geom_point(size=2)+
  #scale_shape_manual(values = c(16, 16))+
  ylim(c(0,1.25))+
  #scale_color_manual(values=c("blue", "red"))+
  labs(x="\nTemperature", y="Variance\n", shape="Temperature", color="Temperature")

g.var
```

Models looking at residuals from beta regression

The first model specifies a random intercept. If mean phagocytosis was the same between groups, the intercept would be the same. 

I then take the residuals from this model and use them as the response variable with groups as the predictor variable, essentially asking, is the variation in the deviation from the overall mean explained by the treatment that the bird was in?

This is still comparing the mean of the residual squares though - should I take the CV of the residuals? 
```{r Residual models: beta regression}
beta_mod1 <- betareg(phago_score ~ 1, data=ti)
ti$resid <-resid(beta_mod1) #residuals
ti$resid2 <-(resid(beta_mod1)^2) #residuals squared
ti$resid3 <- abs(resid(beta_mod1)) #absolute residuals
hist(ti$resid)
#all groups
lm.resid1 <- glm(resid~groups, data=ti)
summary(lm.resid1) #warm infected had more positive residuals = were overall higher than the mean > makes sense
plot(allEffects(lm.resid1))

lm.resid2 <- glm(resid2~groups, data=ti)
summary(lm.resid2) #confused by this one
plot(allEffects(lm.resid2)) #mean residuals were not different between groups

lm.resid3 <- glm(resid3~temp + treatment, data=ti)
summary(lm.resid3) #mean absolute value of residuals was not different between groups
g.ra.effects <- allEffects(lm.resid3) #residuals were not different between groups
plot(g.ra.effects)
g.r <- ggplot(ti, aes(x=groups, y=resid))+
  geom_point()+
  stat_summary(aes(group=groups), fun=mean, color="red")+
  labs(y="Residuals")

g.r2 <- ggplot(ti, aes(x=groups, y=resid2))+
  geom_point()+
  stat_summary(aes(group=groups), fun=mean, color="red")+
  labs(y="Residuals Squared")

g.ra <- ggplot(ti, aes(x=groups, y=resid3))+
  geom_point()+
  stat_summary(aes(group=groups, shape=groups), fun=mean, color="red")+
  labs(y="|Residuals|", x="Treatment Groups", shape="Treatment Groups")
g.r
g.r2
g.ra
```

```{r Residual models: binomial regression}
glm1 <- glmer(phago_score ~ 1 + (1|band_number), weights=wbc_total+phago_total, data=ti, family=binomial)
ti$resid <-resid(glm1) #residuals
ti$resid2 <-(resid(glm1)^2) #residuals squared

#all groups
lm.resid1 <- glm(resid~groups, data=ti)
summary(lm.resid1) #warm infected had more positive residuals = were overall higher than the mean > makes sense
plot(allEffects(lm.resid1))

lm.resid2 <- glm(resid2~groups, data=ti)
summary(lm.resid2) #confused by this one
plot(allEffects(lm.resid2)) #residuals were not different between groups

#temperature
lm.resid3 <- glm(resid2~ temp, data=ti)
summary(lm.resid3)
plot(allEffects(lm.resid3))


ggplot(ti, aes(x=groups, y=resid))+
  geom_point()+
  stat_summary(aes(group=groups), fun=mean, color="red")

ggplot(ti, aes(x=groups, y=resid2))+
  geom_point()+
  stat_summary(aes(group=groups), fun=mean, color="red")
```

Look at the raw data using histograms
```{r Histograms}
treat_names_n <- c(
  'Cold Infected' = "Cold Infected (n=14)",
  'Warm Control' = "Warm Control (n=10)",
  'Warm Infected' = "Warm Infected (n=14)",
  'Cold Control' = "Cold Control (n=10)"
)

#Add variable indicating the phago_score mean for each treatment group
p.ti <- ti%>%
  group_by(groups)%>%
  mutate(groupmean = mean(phago_score))%>%
  ungroup()

m.ti <- p.ti %>%
  mutate(meanall = mean(phago_score)) %>%
  ungroup()

#histogram showing variability in phagocytosis scores across treatment groups
ggplot(p.ti, aes(x=phago_score, fill=groups))+
  geom_histogram(binwidth = 0.01, position="identity", alpha=1, color="black")+
  geom_vline(data = p.ti, aes(xintercept = groupmean, color = "red"), linetype = "dotted",alpha = 1, show.legend=FALSE) + #mean of each group to show differences
  geom_vline(data=m.ti, aes(xintercept = meanall, color = "black"), linetype="dashed", alpha=1, show.legend=FALSE)+ #mean of all groups combined to compare
  scale_fill_manual(values=c( "gray75", "white", "gray35","black"))+
  scale_color_manual(values = c("black", "red"), guide = guide_legend(title= NULL))+
  facet_wrap(~groups, nrow=4, labeller = as_labeller(treat_names_n), scales="fixed")+
  labs(x="Phagocytosis Score", y="Count", fill="Treatment")+
  theme_minimal()
```

I can also calculate coefficient of variation
```{r}
p.ti$N = p.ti$wbc_total + p.ti$phago_total #add column with total number of cells per band_number

m.cv.all <- p.ti %>% 
  group_by(groups)%>%
  summarise(mean.phago = mean(phago_score, na.rm=T), #mean
            all.Ns = sum(N, na.rm = T), #total number of cells
            sd.binom = sqrt(all.Ns*mean.phago*(1-mean.phago)), #standard deviation = square root(N * mean * (1-mean))
            bird_cv = (sd.binom/mean.phago)) #CV = standard deviation / mean * 100

m.cv.all
#generate error bar values
m.cv.all$max <- m.cv.all$bird_cv + m.cv.all$sd.binom
m.cv.all$min <- m.cv.all$bird_cv - m.cv.all$sd.binom
m.cv.all$groups <- factor(m.cv.all$groups, levels = c("Cold Control", "Cold Infected", "Warm Control", "Warm Infected"))
#Graph of CVs
cv.all.primary <- ggplot(m.cv.all, aes(x=groups, y=bird_cv, shape=groups))+
  labs(x="Treatment Groups", y="CV", shape="Treatment Groups", color="Temperature")+
  geom_point(aes(shape=groups), size=3)+
  #scale_shape_manual(values = c(0, 15, 1, 16))+
  #scale_color_manual(values=c("blue", "red"))+
  ylim(c(0, 300))

cv.all.primary
```

*Summary* 
Using a beta regression, I show that infection with MG (p = 0.037), but not temperature at which a bird was held (p = 0.117), significantly affected phagocytic activity during infection. 
```{r}
phago.pred
```

I also show through various methods that the mean variation does not significantly differ across treatments. 

```{r}
g.var #variance from binomial models
cv.all.primary #coefficient of variation 
g.ra #absolute values of residuals
plot(g.ra.effects) #allEffects absolute values of residuals

```

